Declarative Integration of Interactive 3D Graphics into the World-Wide Web:\\Principles, Current Approaches, and Research Agenda

### 2013-04-28 Web3D2013-Dec3D/paper.tex
some polishing

\begin{abstract}
With the advent of WebGL, plugin-free hardware-accelerated interactive 3D graphics has finally arrived in all major Web browsers.
WebGL is an imperative solution that is tied to the functionality of rasterization APIs. Consequently, its usage requires a deeper understanding of the rasterization pipeline. In contrast to this stands a declarative approach with an abstract description of the 3D scene. We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related Web technologies, as those concepts are well-known by millions of Web developers and therefore crucial for the fast adoption of 3D on the Web.
Hence, in this paper we explore the options for new declarative ways of incorporating 3D graphics directly into HTML to enable its use on any Web page. We present declarative 3D principles that guide the work of the \textit{Declarative 3D for the Web Architecture W3C Community Group} and describe the current state of the fundamentals to this initiative. Finally, we draw an agenda for the next development stages of Declarative 3D for the Web.
\end{abstract}

@@

It is arguable that the emerging support for an imperative 3D API for the Web is useful but insufficient for broad acceptance and usage of 3D on the Web. A declarative approach that offers qualified concepts and that is tightly integrated with current web technologies, such as JSON or XML for scene construction, DOM for scene manipulation, and CSS for styling, is necessary to support a fast adoption and broad use of interactive 3D graphics by the millions of Web developers. The provided concepts must lift the hardware-oriented imperative application programming interfaces to an expressive and more easily usable level. Therefore not the low-level data structures of existing GPU layers must be in the center of the design but high-level elements and items like 3D objects, transformations, material descriptions, and lights. Instead of teaching Web developers 3D graphics APIs, the goal is to bring 3D graphics to the point where it is natural for Web developers to just make use of it. While this might not be possible for every possible use of a low-level API, it still can cover the vast majority of use cases.

The \textit{Declarative 3D for the Web Architecture W3C Community Group} (Dec3D) \cite{Dec3D} was thus founded to suggest and create methods to add high-level declarative 3D objects to the HTML DOM \cite{W3C-DOM}, so users can share and experience interactive 3D contents. Moreover, this not only allows creating new content from existing content but also to index and search 3D content.
The core mission is to determine the requirements, options, and use cases for the integration of interactive 3D graphics into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
Therefore, the group aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality, Efficient XML Interchange (EXI) \cite{EXI} and other key working groups, whereas certain complex data types (e.g., transformation matrices) and computations are also of mutual interest. In this regard, this paper presents the current state and efforts concerning Dec3D.

@@

\paragraph{Platform Independence}
3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only \cite{Rubinstein:2009,Schwenk12}. This should allow for content to be portable across web browsers, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible. The results of rendering content under such different environments should be highly predictable.

@@

Fraunhofer IGD's X3DOM \cite{Behr2009,Behr12} is a JavaScript-based open-source framework for declarative 3D graphics in HTML5 that aims at extending the HTML DOM tree with declarative 3D objects while employing modern Web technologies like CSS3, Ajax, DOM scripting, as well as WebGL and -- as fallback -- Adobe's Flash 11 with Stage 3D \cite{flash} for GPU-accelerated rendering. The proposed 3D elements are mostly based upon the open ISO standard X3D \cite{Web3D-X3D}, though X3DOM introduces a special HTML profile that basically extends the X3D Interchange profile. Additionally, instead of implementing the somehow odd X3D pointing device sensor component, X3DOM simply uses, and appropriately extends, the HTML UI/Mouse events such that 3D pick events are likewise supported.

@@

It is also worth mentioning that the Web applications shown in Figures \ref{fig:chApps} and \ref{fig:cadApps} were not realized by graphics programmers but mainly by Web developers, using standard DOM scripting as well as JavaScript frameworks like jQuery \cite{jquery}. For example, to switch to a certain viewpoint when the user clicks onto the arrow button in the walkthrough application shown in Figure \ref{fig:chApps} (left), the application developer simply needs to update the 'bind' attribute of the "viewpoint" tag by calling its \emph{setAttribute} method.

XML3D \cite{Sons2010} is a collaboration of DFKI and Intel VCI. Similar to X3DOM, it employs CSS3, DOM scripting, DOM events, and WebGL. In contrast to X3DOM, XML3D is designed as a extension to HTML5 from scratch. It does not define a scene-graph but uses the DOM tree structure for parent child relations and second level references e.g\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data. These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g. as JSON. % files.

This generic and consistent handling of resources allows fine-granular composition of data sources. It can be combined with Xflow, a declarative data flow component~\cite{KleinSJRSB12} that allows for dynamic meshes, morphing, animation of shader parameters, image processing and AR. The data flow computation can be mapped to hardware, e.g. by composition of WebGL shaders or by exploiting available APIs such as Web Workers~\cite{w3c:webworkers} or River Trail~\cite{Herhut:2012}. Figure~\ref{fig:xflow} shows a XML3D scene with nine characters, skinned and animated with Xflow.

@@

The CG identified 15 key concepts of DOM-based Dec3D. In this section we briefly outline our proposed essential elements for an upcoming standard.

@@

In general, a tight \textit{CSS integration}~(4) is a major concept: Existing CSS properties should be used where applicable (e.g.\ transform, opacity, color, etc.). In addition, a set of 3D-specific CSS properties (e.g. for defining the appearance) needs to be defined. However, custom CSS properties are not yet supported by any Web Browser, though finally this seems to change now \cite{CSSCustom}.

For realtime interactive 3D graphics systems an \textit{event}~(5) model is essential. The Dec3D standard should leverage the DOM Event model where applicable. Not only standard DOM UI Events such as `mouseover' and `click' shall be supported, but also new events need to be defined to provide 3D-specific context information. In this regard, the Browser API should expose convenience functions and special Dec3D \textit{base types}~(6) consisting of a lean set of complex data types for 2D/3D graphics. In general, we aim a common set of base types usable not only for Dec3D, but also for WebGL API, Audio API, SVG, 3D Transforms etc.

@@

Dec3D should support highly dynamic scenes. Using the DOM API to modify the scene data is not sufficient for many use-cases, e.g.\ dynamic meshes, image processing, compression, etc. Thus we need a concept to support efficient data processing that allows to map computations to available hardware. These should be expressed as \textit{operators}~(10) on the data containers above.

@@

Essential DOM elements are a basic set of pre-defined \textit{shaders}~(11), which should be enough for most use cases. Advantages are that they do not entail timing attack issues and that they can adapt to any target device. However, the design should not obstruct programmable shaders in the future. HTML elements such as \verb|<img>|, \verb|<video>|, and \verb|<canvas>| can be used as \textit{textures}~(12), whereas elements such as \verb|<svg>| and \verb|<html>| could serve as interactive textures -- once the corresponding security issues are resolved.

\textit{Lights}~(13) are shader parameters in OpenGL/WebGL \cite{WebGL12}. In contrast, for Dec3D we aim a representation to provide a intuitive way to define lights and to not obstruct more advanced future concepts.
Likewise, \textit{viewpoints}~(14) can serve as link to certain points of interest within the document. However, it is still discussable if some pre-defined \textit{camera navigation}~(15) modes should be part of the essential profile.

@@

% Don't cite this link but present the whole idea here!
% http://www.w3.org/community/declarative3d/wiki/Level_of_Integration/
We also propose to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS, etc.) to explain existing and possible integration levels (Figure \ref{fig:polyInt} shows that currently only level 2, DOM integration, is possible), and to finally identify a suitable integration level for declarative 3D. However, the latter imposes several problems in that it requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug. However, integration levels 3 and 4 are currently not yet achievable by a JavaScript implementation, because the required APIs are missing: it is not possible to introduce new CSS properties or to extend developer tools by an API.

It quickly became clear that a higher level of integration requires additional APIs in user agents. After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a polyfill approach for the further development activities.
Nevertheless, communicating and requesting additional and missing features like custom CSS properties or CSS monitoring is necessary.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.85\columnwidth]{images/polyfill.png}
  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
  \label{fig:polyInt}
  %\vspace{-0.2cm}
\end{figure}

@@

\paragraph{Clear Definition of Use Cases and Requirements}
The group needs to agree on a collection of use cases, where embedding 3D data in HTML using a declarative approach provides significant benefit. Here, declarative approaches esp. seem relevant for the industry, since they tend to think in formats not in APIs.
Each use case should explore how publishers and consumers benefit from Dec3D. From these use cases, the group needs to derive and prioritize the different required dimensions for the Dec3D technical specification \cite{JankowskiDec3D2012,LeFeuvreDec3D2012}.

@@

\paragraph{Outreach and Exemplar Applications}
The group needs to continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the open-source frameworks X3DOM and XML3D. Therefore, several applications should be identified, each requiring and demonstrating different capabilities of Dec3D. For example, one application could require huge 3D datasets which are impractical for inclusion directly into the DOM; another could require real-time control of an external system (inter-process communication); another could integrate with a complex data base; another an illustration of complex data (information visualization) without a physical analog, and another a more traditional scientific visualization. Key to selection of these applications is the use and demonstration of a declarative 3D requirement.

@@

While WebGL, a 3D imperative graphics API in the Web context, is getting more and more traction, we are still missing an easy way to add interactive high-level declarative 3D objects to an HTML document to allow anyone to easily create, share, and experience interactive 3D graphics, with possibly wide ranging effects similar to those caused by the broad availability of video on the Web.
The main motivation is thus to make it easy to add 3D to Web pages by bringing 3D to the Web developers and not vice versa. This can be achieved by fully integrating 3D content into HTML5 documents, where interactive 3D graphics is a first class DOM objects. Moreover, by reusing existing Web technology wherever possible, no new concepts are added unless absolutely necessary.

>>>>

\begin{abstract}
With the advent of WebGL, plugin-free hardware-accelerated interactive 3D graphics has finally arrived in all major Web browsers.
WebGL is an imperative solution that is tied to the functionality of rasterization APIs. Consequently, its usage requires a deeper understanding of the rasterization pipeline. In contrast to this stands a declarative approach with an abstract description of the 3D scene. We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related Web technologies, as those concepts are well-known by millions of Web developers and therefore crucial for the fast adoption of 3D on the Web.
Hence, in this paper we explore the options for new declarative ways of incorporating 3D graphics directly into HTML to enable its use on any Web page. We present declarative 3D principles that guide the work of the \emph{Declarative 3D for the Web Architecture W3C Community Group} and describe the current state of the fundamentals to this initiative. Finally, we draw an agenda for the next development stages of Declarative 3D for the Web.
\end{abstract}

@@

It is arguable that the emerging support for an imperative 3D API for the Web is useful but insufficient for broad acceptance and usage of 3D on the Web. A declarative approach that offers qualified concepts and that is tightly integrated with current web technologies, such as JSON or XML for scene construction, DOM for scene manipulation, and CSS for styling, is necessary to support a fast adoption and broad use of interactive 3D graphics by the millions of Web developers. The provided concepts must lift the hardware-oriented imperative application programming interfaces to an expressive and more easily usable level. Therefore, not the low-level data structures of existing GPU layers must be in the center of the design but high-level elements and items like 3D objects, transformations, material descriptions, and lights. Instead of teaching Web developers 3D graphics APIs, the goal is to bring 3D graphics to the point where it is natural for Web developers to just make use of it. While this might not be feasible for every possible use of a low-level API, it still can cover the vast majority of use cases.

The \textit{Declarative 3D for the Web Architecture W3C Community Group} (Dec3D) \cite{Dec3D} was thus founded to suggest and create methods to add high-level declarative 3D objects to the HTML DOM \cite{W3C-DOM}, so users can share and experience interactive 3D contents. Moreover, this not only allows creating new content from existing content but also to index and search 3D content.
The core mission is to determine the requirements, options, and use cases for the integration of interactive 3D graphics into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.

The group thus aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality, Efficient XML Interchange (EXI) \cite{EXI} and other key working groups, whereas certain complex data types (e.g., transformation matrices) and computations are also of mutual interest. In this regard, this paper presents the current state and efforts concerning Dec3D.

@@

\paragraph{Platform Independence}
3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only (cp. e.g. \cite{Rubinstein:2009,Schwenk12}). This should allow for content to be portable across web browsers, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible. The results of rendering content under such different environments should be highly predictable.

@@

Fraunhofer IGD's \emph{X3DOM} \cite{Behr2009,Behr12} is a JavaScript-based open-source framework for declarative 3D graphics in HTML5 that aims at extending the HTML DOM tree with declarative 3D objects while employing modern Web technologies like CSS3, Ajax, DOM scripting, as well as WebGL and -- as fallback -- Adobe's Flash 11 with Stage 3D \cite{flash} for GPU-accelerated rendering. The proposed 3D elements are mostly based upon the open ISO standard X3D \cite{Web3D-X3D}, though X3DOM introduces a special HTML profile that basically extends the X3D Interchange profile. Additionally, instead of implementing the somehow odd X3D pointing device sensor component, X3DOM simply uses, and appropriately extends, the HTML UI/Mouse events such that 3D pick events are likewise supported.

@@

It is also worth mentioning that the Web applications shown in Figures \ref{fig:chApps} and \ref{fig:cadApps} were not realized by graphics programmers but mainly by Web developers, using standard DOM scripting as well as JavaScript frameworks like jQuery \cite{jquery}. For example, to switch to a certain viewpoint when the user clicks onto the arrow button in the walkthrough application shown in Figure \ref{fig:chApps} (left), the application developer simply needs to update the 'bind' attribute of the "viewpoint" tag by calling its \emph{setAttribute} method.
Likewise, when selecting an element in the browser-based CAD and product structure viewer shown in the middle of Figure \ref{fig:cadApps}, the corresponding parts in the tree view are highlighted and vice versa.

\emph{XML3D} \cite{Sons2010} is a collaboration of DFKI and Intel VCI. Similar to X3DOM, it employs CSS3, DOM scripting, DOM events, and WebGL. In contrast to X3DOM, XML3D is designed as a extension to HTML5 from scratch. It does not define a scene-graph but uses the DOM tree structure for parent child relations and second level references e.g\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data. These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g. as JSON.

This generic and consistent handling of resources allows fine-granular composition of data sources. It can be combined with Xflow, a declarative data flow component~\cite{KleinSJRSB12} that allows for dynamic meshes, morphing, animation of shader parameters, image processing and Augmented Reality. The data flow computation can be mapped to hardware, e.g. by composition of WebGL shaders or by exploiting available APIs such as Web Workers~\cite{w3c:webworkers} or River Trail~\cite{Herhut:2012}. Figure~\ref{fig:xflow} shows a XML3D scene with nine characters, skinned and animated with Xflow.

@@

The community group identified 15 key concepts of DOM-based Dec3D. In this section we briefly outline our proposed essential elements for an upcoming standard.

@@

In general, a tight \textit{CSS integration}~(4) is a major concept: existing CSS properties should be used where applicable (e.g., transform, opacity, color, etc.). In addition, a set of 3D-specific CSS properties (e.g. for defining the appearance) needs to be defined. However, custom CSS properties are not yet supported by any Web Browser, though finally this seems to change now \cite{CSSCustom}.

For real-time interactive 3D graphics systems an \textit{event}~(5) model is essential. The Dec3D standard should leverage the DOM Event model where applicable. Not only standard DOM UI Events such as `mouseover' and `click' (with appropriate extensions for 3D scenes as already discussed in \cite{BehrJDA11}) shall be supported, but also new events need to be defined to provide 3D-specific context information. In this regard, the Browser API should expose convenience functions and special Dec3D \textit{base types}~(6) consisting of a lean set of complex data types for 2D/3D graphics. In general, we aim a common set of base types usable not only for Dec3D, but also for WebGL API, Audio API, SVG, 3D Transforms etc.

@@

Dec3D should support highly dynamic scenes. Using the DOM API to modify the scene data is not sufficient for many use cases, e.g.\ dynamic meshes, image processing, compression, etc. Thus we need a concept to support efficient data processing that allows to map computations to available hardware. These should be expressed as \textit{operators}~(10) on the data containers above.

@@

Essential DOM elements include a basic set of pre-defined \textit{shaders}~(11), which should be enough for typical use cases. Advantages are that they do not entail timing attack issues and that they can adapt to any target device. However, the design should not obstruct programmable shaders in the future. HTML elements such as \verb|<img>|, \verb|<video>|, and \verb|<canvas>| can be used as \textit{textures}~(12), whereas elements such as \verb|<svg>| and \verb|<html>| could serve as interactive textures -- once the corresponding security issues are resolved.

\textit{Lights}~(13) are shader parameters in OpenGL/ WebGL \cite{WebGL12}. In contrast, for Dec3D we aim a representation to provide a intuitive way to define lights (and thereby also shadows) and to not obstruct more advanced future concepts.
Likewise, \textit{viewpoints}~(14) can serve as link to certain points of interest within the document. However, it is still discussable if some pre-defined \textit{camera navigation}~(15) modes -- to get directly started without having to modify the camera matrix with a special script -- should be part of the essential profile.

@@

% http://www.w3.org/community/declarative3d/wiki/Level_of_Integration/
We also propose to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS, etc.) to explain existing and possible integration levels (Figure \ref{fig:polyInt} shows that currently only level 2, DOM integration, is possible here), and to finally identify a suitable integration level for declarative 3D. The latter imposes several problems in that it requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug. However, integration levels 3 and 4 are currently not yet achievable by a JavaScript implementation, because the required APIs are missing: it is not possible to introduce new CSS properties or to extend developer tools by an API.

It quickly became clear that a higher level of integration requires additional APIs in user agents. After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a so-called polyfill approach for the further development activities.
Nevertheless, communicating and requesting additional and missing features like custom CSS properties or CSS monitoring is necessary.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\columnwidth]{images/polyfill.png}
  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
  \label{fig:polyInt}
  %\vspace{-0.2cm}
\end{figure}

@@

\paragraph{Clear Definition of Use Cases and Requirements}
The group needs to agree on a collection of relevant use cases, where embedding 3D data in HTML using a declarative approach provides significant benefit. Here, declarative approaches esp. seem relevant for the industry, since they tend to think in formats (such as STL, STEP/IGES or CATIA) not in APIs.
Each use case should explore how publishers and consumers benefit from Dec3D. From these use cases, the group needs to derive and prioritize the different required dimensions for the Dec3D technical specification \cite{JankowskiDec3D2012,LeFeuvreDec3D2012}.

@@

\paragraph{Outreach and Exemplar Applications}
The group needs to further continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the open-source frameworks X3DOM and XML3D. Therefore, several applications have to be identified, each requiring and demonstrating different capabilities of Dec3D. For example, one application could require huge 3D datasets, which are impractical for inclusion directly into the DOM; another could require real-time control of an external system; another could integrate with a complex data base; another an illustration of complex data (information visualization) without a physical analog, and yet another a more traditional scientific visualization. Key to selection of these applications is the use and demonstration of a more declarative 3D requirement.

@@

While WebGL, a 3D imperative graphics API in the Web context, is getting more and more traction, we are still missing an easy way to add interactive high-level declarative 3D objects to an HTML document to allow anyone to easily create, share, and experience interactive 3D graphics, with possibly wide ranging effects similar to those caused by the broad availability of video on the Web.
The main motivation is thus to make it easy to add 3D graphics to Web pages by bringing 3D to the Web developers and not vice versa. This can be achieved by fully integrating 3D content into HTML5 documents, where interactive 3D graphics is a first class DOM objects. Moreover, by reusing existing Web technology wherever possible, no new concepts are added unless absolutely necessary.
